\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-07 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
update competing risk

In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked how he would model a set of censored survival data, to which he responded: "I think I would normally want to tackle problems parametrically ... and if you want to do things like predict the outcome for a particular patient, it's much more convenient to do that parametrically" [@reid1994conversation]. Indeed, the most relevant quantity is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances. Unfortunately, the most reported metric from a Cox model is the (potentially time-dependent) hazard ratio (HR), which ignores the duration of follow-up or is subject to selection bias [@hernan2010hazards]. The covariate adjusted survival curve overcomes these limitations, and is arguably a more important summary measure to report than the HR. Stepwise survival curves can be computed with the Cox model but require a second step to separately estimate the baseline hazard [@breslow1972discussion].

In this article, we present the \CRANpkg{casebase} R package [@casebase-package] for fitting fully parametric hazard models and covariate adjusted survival curves using the familiar interface of the \code{glm} function. Using a parametric estimator of the hazard function leads to a smooth function of time; as a consequence, the survival function estimates also vary smoothly over time.

Other packages can do this.

the familiar interface of the \code{glm} function. It also provides tools for variable selection, competing-risk analysis, and dealing with time-dependent exposures. In addition, we provide functions for exploratory data analysis and visualizing the estimated quantities.

@@

For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms. We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor (Eq. \eqref{eq:exp}); 2) linear function (Eq. \eqref{eq:gomp}); 3) log function (Eq. \eqref{eq:weibull}); and 4) a smooth function using cubic B-splines. The models are fit using \code{fitSmoothHazard} with the familiar formula interface:

```{r, eval = eval_cs1, echo = -1}
set.seed(12)
fmla <- list(exponential = formula(DeadOfPrCa ~ ScrArm),
             gompertz = formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
             weibull = formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
             splines = formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))

fits <- lapply(fmla, function(form) {
  fitSmoothHazard(form, data = ERSPC, ratio = 100)
})
```

The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes. For this reason, we leveraged the \code{summary} method for \code{glm} objects to output a familiar summary of the results:

```{r, eval = TRUE, echo=eval_cs1}
summary(fits[["splines"]]) 
```

@@

```{r, eval=FALSE, echo=echo_plot_code}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0, 14, by = 0.1)

risks <- lapply(fits, function(fit) {
  absoluteRisk(fit, time = new_time, newdata = new_data)
})
```

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \pkg{casebase} on the Cox model CIF. The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on. Based on our earlier discussion of the population-time plot, this poor fit for the exponential hazard was expected. We notice a better fit with increasing complexity of our model for time in Figure \ref{fig:erspc-cox-cif} (panels 2--4). As noted in Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) spline model:

```{r erscp-compare, eval = eval_cs1, echo = FALSE}
anova(fits$exponential, fits$splines, test = "LRT")
```

The null hypothesis is rejected in favor of the spline model. Similarly, the AIC provides further evidence that the flexible function of time provides the best fit:

```{r erscp-aic, eval = eval_cs1, echo=FALSE}
c("Exp." = AIC(fits$exponential),
  "Gompertz" = AIC(fits$gompertz),
  "Weibull" = AIC(fits$weibull),
  "Splines" = AIC(fits$splines))
```

@@

```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIFs for control and screening groups in the ERSPC data. In each of the panels, we plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time. (1) The time variable is excluded (exponential). (2) Linear function of time (Gompertz). (3) The natural logarithm (Weibull). (4) Cubic B-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]

new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14.5,0.1)

risks <- lapply(fits, function(i) {
  casebase::absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
})

layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
par(mar = c(0, 4.1, 4.1, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",
     xlab = "",
     fun = "event",
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Case-base)",
                  "Screening (Cox)", "Screening (Case-base)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(1)", cex = 1.5)

par(mar = c(4.1, 4.1, 0, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "", xlab = "",
     fun = "event",
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$weibull[,1], risks$weibull[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(3)", cex = 1.5)
mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)

par(mar = c(0, 0, 4.1, 3.9))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",xlab = "",
     fun = "event",
      xaxt = 'n', yaxt = 'n',
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$gompertz[,1], risks$gompertz[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(2)", cex = 1.5)

par(mar = c(4.1, 0, 0, 3.9))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",xlab = "",
     fun = "event",
     yaxt = 'n',
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$splines[,1], risks$splines[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(4)", cex = 1.5)
mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
```

```{r erspc-95ci22, echo = FALSE, eval = eval_cs1}
CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))

aft_fits <- lapply(list("exp","gompertz","weibull"), function(i)
  flexsurv::flexsurvreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
                           data = ERSPC, dist = i)
  )

names(aft_fits) <- list("exponential","gompertz","weibull")

# function to exponentiate HR and CI and return a HR (lower, Upper) character
exp_est_ci <- function(model = c("exponential","gompertz","weibull","splines"),
                       fit, 
                       covariate = "ScrArmScreening group", 
                       type = c("casebase", "survreg")) {

  type <- match.arg(type)
  covariate <- match.arg(covariate)

  if (type == "casebase"){

    cis <- confint(fit[[model]])

    sprintf("%0.2f (%0.2f, %0.2f)",
            exp(coef(fit[[model]]))[[covariate]],
            exp(cis)[covariate,1],
            exp(cis)[covariate,2])
  } else {

    if (model == "weibull") {

      # The maximized log-likelihoods are the same between survreg and flexsurvreg, 
      # however the parameterisation is different: the first coefficient (Intercept) 
      # reported by survreg is log(mu), and survreg's "scale" is
      # dweibull's (thus flexsurvreg)'s 1 / shape
      res <- -fit[[model]]$res[covariate,] / (1 / fit[[model]]$res["shape",])

      sprintf("%0.2f (%0.2f, %0.2f)",
              exp(res[["est"]]),
              exp(res[["U95%"]]),
              exp(res[["L95%"]]))

    } else if (model == "splines") {

      c("--")

    } else {

      res <- fit[[model]]$res[covariate,]

      sprintf("%0.2f (%0.2f, %0.2f)",
              exp(res[["est"]]),
              exp(res[["L95%"]]),
              exp(res[["U95%"]]))
    }
  }
}

casebase_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
                         exp_est_ci, 
                         fit = fits,
                         covariate = "ScrArmScreening group", 
                         type = "casebase")

survreg_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
                         exp_est_ci, 
                         fit = aft_fits,
                         covariate = "ScrArmScreening group", 
                         type = "survreg")

ptable <- cbind(list("Exponential","Gompertz","Weibull","Splines"),
            do.call(rbind, casebase_HR_CI), do.call(rbind, survreg_HR_CI))
```

In Table \ref{tab:print-erspc-estimates}, we present a side-by-side comparison of the hazard ratios and confidence intervals estimated from \code{fitSmoothHazard} and the corresponding parametric model using \code{survival::survreg}, as well as the Cox model estimate. The hazard ratio estimates and confidence intervals are similar across all four models. This reinforces the idea that, under proportional hazards, we do not need to model the full hazard to obtain reliable estimates of the hazard ratio. Nevertheless, Figure \ref{fig:erspc-cox-cif} shows that different parametric models can still give rise to qualitatively different estimates for the CIF.

```{r print-erspc-estimates, eval = eval_cs1}
kable(ptable, "latex", booktabs = TRUE, align = c("l","c","c"),
      caption = c("Comparison of estimated hazard ratios and 95\\% confidence intervals for ERSPC data."),
      col.names = c("Model","casebase::fitSmoothHazard","survival::survreg")) %>%
  kable_styling() %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1, latex_hline = "major", valign = "middle") %>%
  footnote(general = sprintf("Cox model estimate: HR (95%% CI) = %0.2f (%0.2f, %0.2f)", CI_cox[1], CI_cox[2], CI_cox[3]),
           general_title = "") 
```

@@

We want to compare our hazard ratio estimates to that obtained from a Cox regression. To obtain these estimates, we need to treat the competing event as censoring:

```{r bmtcrr-cox, echo = TRUE, eval = eval_cs2}
library(survival)
# Treat competing event as censoring
model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
                   data = bmtcrr)
```

@@

table_cox <- round(summary(model_cox)$conf.int[, -2], 2)
rownames(table_cb) <- rownames(table_cox)
colnames(table_cb) <- colnames(table_cox)

@@

Given the estimate of the hazard function obtained using case-base sampling, we can compute the absolute risk curve for a fixed covariate profile. We perform this computation for a 35 year old woman who received a stem-cell transplant from peripheral blood at relapse. We compare the absolute risk curve for such a woman with ALL with that for a similar woman with AML.

@@

# Estimate absolute risk curve
risk_fg <- predict(model_fg, newdata, times = time_points)
```

```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.", eval = eval_cs2}
risk_all <- dplyr::bind_rows(
  data.frame(
    Time = time_points,
    Method = "Case-base",
    Risk = risk_cb[, 2],
    Disease = "ALL",
    stringsAsFactors = FALSE
  ),
  data.frame(
    Time = time_points,
    Method = "Case-base",
    Risk = risk_cb[, 3],
    Disease = "AML",
    stringsAsFactors = FALSE
  ),
  data.frame(
    Time = time_points,
    Method = "Fine-Gray",
    Risk = risk_fg$P1[1, ],
    Disease = "ALL",
    stringsAsFactors = FALSE
  ),
  data.frame(
    Time = time_points,
    Method = "Fine-Gray",
    Risk = risk_fg$P1[2, ],
    Disease = "AML",
    stringsAsFactors = FALSE
  )
)

@@

Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both case-base sampling and the Fine-Gray model. As we can see, the two approaches agree quite well for AML; however, there seems to be a difference of about 5% between the two curves for ALL. This difference does not appear to be significant: the curve from case-base sampling is contained within a 95% confidence band around the Fine-Gray absolute risk curve (figure not shown).

@@

In this article, we presented the R package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling. Our package also provides several functions to produce graphical summaries of the data and the results. We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.

@@

As presented in Hanley & Miettinen [-@hanley2009fitting], case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the CIF. Accordingly, our package provides functions for each step. Moreover, the simple interface of the \code{fittingSmoothHazard} function resembles the \code{glm} interface. This interface should look familiar to new users. Our modular approach also provides a convenient way to extend our package for new sampling or fitting strategies.

In the case studies above, we compared the performance of case-base sampling with that of Cox regression and Fine-Gray models. In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula. By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that capture both the event type and time. Both approaches to modeling yield user-friendly code. However, in terms of output, both approaches differ significantly. Case-base sampling produces smooth hazards and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce stepwise CIFs and never explicitly model the hazard function. Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves. However, the smooth nature of the output of \pkg{casebase} provides a more intuitive interpretation for consumers of these predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling.

\begin{widetable}
\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
\centering
\begin{tabular}[t]{llp{5cm}}
\toprule
Feature & Cox model & Case-base sampling\\
\toprule
Model type & Semi-parametric & Fully parametric\\\midrule
Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\\midrule
Cumulative incidence & Step function & Smooth-in-time curve\\\midrule
Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\\midrule
Model testing & Likelihood-ratio tests, score tests, & Use GLM framework\\
 & information criteria & (e.g.\ LRT, AIC, BIC)\\\midrule
\addlinespace
Competing risks & Difficult & Cause-specific CIFs\\
% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
\bottomrule
\end{tabular}
\end{widetable}

@@

In future versions of the package, we also want to increase the complement of diagnostic and inferential tools that are currently available. For example, we would like to include the ability to compute confidence intervals for the cumulative incidence curve. The delta method or parametric bootstrap are two different strategies we can use to construct approximate confidence intervals. Furthermore, we would like to include more functions to compute calibration and discrimination statistics (e.g. AUC) for our models. Saarela and Arjas [-@saarela2015non] also describe how to obtain a posterior distribution for the AUC from their model. Their approach could potentially be included in \pkg{casebase}. Finally, we want to provide more flexibility in how the case-base sampling is performed. This could be achieved by adding a \code{hazard} argument to the function \code{sampleCaseBase}. In this way, users could specify their own sampling mechanism. For example, they could provide a hazard that gives sampling probabilities that are proportional to the cardiovascular disease event rate given by the Framingham score [@saarela2015non].

In conclusion, we presented the R package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth cumulative incidence functions using the framework of generalized linear models. We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practitioners. The \pkg{casebase} package is freely available from the Comprehensive R Archive Network at \url{https://cran.r-project.org/package=casebase}. Interested users can visit \url{http://sahirbhatnagar.com/casebase/} for detailed package documentation and vignettes.

>>>>

@@

For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms. We fit a flexible hazard by using a smooth function of time modeled with cubic B-splines. The model is fit using \code{fitSmoothHazard} with the familiar formula interface:

```{r, eval = eval_cs1, echo = -1}
set.seed(12)
library(splines)
fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
                       data = ERSPC, ratio = 100)
```

The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes. For this reason, we can leverage the \code{summary} method for \code{glm} objects to output a familiar summary of the results:

```{r, eval = eval_cs1, echo = TRUE}
summary(fit) 
```

@@

```{r, eval=eval_cs1, echo=echo_plot_code}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- c(0, 1)

absoluteRisk(fit, time = new_time, newdata = new_data)
```

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk function from \pkg{casebase} on the Cox model estimate. As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. Therefore, we can easily test the significance of the spline model:

```{r erscp-compare, eval = eval_cs1, echo = FALSE}
anova(fit, test = "LRT")
```

The null hypothesis is rejected in favor of the spline model. Similarly, to compare different models (e.g. time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.

@@

```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data. We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line). We used a cubic B-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]

new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
# Recompute with more time points
risk_mp <- absoluteRisk(fit, time = new_time, newdata = new_data)

plot(survfit(cox_model, newdata = new_data),
     ylab = "Risk probability (%)",
     xlab = "Years since randomization",
     fun = "event",
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Case-base)",
                  "Screening (Cox)", "Screening (Case-base)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risk_mp[,1], risk_mp[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risk_mp[,1], risk_mp[,3], col = myCols[2], lty = 2, lwd = 2)
```

Moreover, we can also compute confidence intervals for the risk function using the method \code{confint.absRiskCB}.

```{r eval = eval_cs1, echo = -1}
new_time <- c(0, 1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
conf_ints <- confint(risk, fit)
conf_ints
```

These point-wise confidence intervals are constructed using parametric bootstrap; the validity of this approach relies on the asymptotic normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates.

```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
# confidence interval
tt <- confint(risk_mp, fit)

# plot absrisk and 95% CI
new_time <- seq(0,14,0.1)
i.backw <- order(new_time, decreasing = TRUE)
i.forw <- order(new_time)

x.poly <- c(new_time[i.forw], new_time[i.backw])
range_y <- 100 * range(c(tt$conf.high, tt$conf.low)) * c(0.99, 1.01)

par(mfrow = c(1, 2))
# 1. Control
tt_sub <- subset(tt, cov_prof == "Control group")
y.poly <- 100 * c(tt_sub$conf.low[i.forw], 
                  tt_sub$conf.high[i.backw])
do.call("plot", 
    list(
        x = range(x.poly),
        y = range_y,
        type = "n",
        ylab = "Risk probability (%)",
        xlab = "Years since randomization",
        main = "Control group"
    )
)

graphics::polygon(x.poly, y.poly, col = "lightgray", border = NA)

# Add KM
km_fit <- survfit(Surv(Follow.Up.Time, DeadOfPrCa) ~ 1,
                  data = ERSPC,
                  subset = ScrArm == "Control group",
                  conf.type = "log-log")

lines(x = km_fit$time, y = 100 * (1 - km_fit$surv),
      lwd = 1, lty = 1, col = myCols[2])
lines(x = km_fit$time, y = 100 * (1 - km_fit$lower),
      lwd = 2, lty = 3, col = myCols[2])
lines(x = km_fit$time, y = 100 * (1 - km_fit$upper),
      lwd = 2, lty = 3, col = myCols[2])

do.call("lines", 
        list(
            x = new_time[i.forw],
            y = 100 * tt_sub$estimate[i.forw],
            lwd = 2,
            lty = 1,
            col = myCols[1]
        )
)

# Add legend
legend("topleft", 
       legend = c("Case-base", "Kaplan-Meier"), 
       col = myCols,
       lty = c(1, 1), 
       lwd = c(2, 1),
       cex = 1.1,
       bg = "gray90")

# 2. Screening
tt_sub <- subset(tt, cov_prof == "Screening group")
y.poly <- 100 * c(tt_sub$conf.low[i.forw], 
                  tt_sub$conf.high[i.backw])
do.call("plot", 
    list(
        x = range(x.poly),
        y = range_y,
        type = "n",
        # ylab = "Risk probability (%)",
        ylab = "",
        xlab = "Years since randomization",
        main = "Screening group"
    )
)

graphics::polygon(x.poly, y.poly, col = "lightgray", border = NA)

# Add KM
km_fit <- survfit(Surv(Follow.Up.Time, DeadOfPrCa) ~ 1,
                  data = ERSPC,
                  subset = ScrArm == "Screening group",
                  conf.type = "log-log")

lines(x = km_fit$time, y = 100 * (1 - km_fit$surv),
      lwd = 1, lty = 1, col = myCols[2])
lines(x = km_fit$time, y = 100 * (1 - km_fit$lower),
      lwd = 2, lty = 3, col = myCols[2])
lines(x = km_fit$time, y = 100 * (1 - km_fit$upper),
      lwd = 2, lty = 3, col = myCols[2])

do.call("lines", 
        list(
            x = new_time[i.forw],
            y = 100 * tt_sub$estimate[i.forw],
            lwd = 2,
            lty = 1,
            col = myCols[1]
        )
)
```

@@

We want to compare our hazard ratio estimates to that obtained from a Cox regression.

```{r bmtcrr-cox, echo = TRUE, eval = eval_cs2}
library(survival)
# Prepare data for coxph
bmtcrr_cox <- transform(bmtcrr, 
                        id = seq_len(nrow(bmtcrr)),
                        Status = factor(Status))

model_cox <- coxph(Surv(ftime, Status) ~ Sex + D + Phase + Source + Age,
                   data = bmtcrr_cox, id = id)
```

@@

# Identify relevant coefficients
coef_relev <- grepl("_1:2$", names(coef(model_cox)))
table_cox <- round(summary(model_cox)$conf.int[coef_relev, -2], 2)
rownames(table_cox) <- gsub("_1:2$", "", rownames(table_cox))
rownames(table_cb) <- rownames(table_cox)
colnames(table_cb) <- colnames(table_cox)

@@

Given the estimate of the hazard functions obtained using case-base sampling, we can compute the cumulative incidence curve for a fixed covariate profile. We perform this computation for a 35 year old woman who received a stem-cell transplant from peripheral blood at relapse. We compare the absolute risk curve for such a woman with ALL with that for a similar woman with AML.

@@

# Estimate CI curve
risk_fg <- predict(model_fg, newdata, times = time_points)
```

We can also estimate the CI for relapse using the Cox model and the Aalen-Johansen estimator:

```{r cox_risk, eval = eval_cs2, echo = TRUE}
# Estimate absolute risk curve
risk_cox <- survfit(model_cox, newdata = newdata)
```

```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk} Cumulative Incidence curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.", eval = eval_cs2}
risk_all <- dplyr::bind_rows(
  data.frame(
    Time = time_points,
    Method = "Case-base",
    Risk = risk_cb[, 2],
    Disease = "ALL",
    stringsAsFactors = FALSE
  ),
  data.frame(
    Time = time_points,
    Method = "Case-base",
    Risk = risk_cb[, 3],
    Disease = "AML",
    stringsAsFactors = FALSE
  ),
  data.frame(
    Time = risk_cox$time,
    Method = "Cox",
    Risk = risk_cox[,2]$pstate[,1,1],
    Disease = "ALL",
    stringsAsFactors = FALSE
  ),
  data.frame(
    Time = risk_cox$time,
    Method = "Cox",
    Risk = risk_cox[,2]$pstate[,2,1],
    Disease = "AML",
    stringsAsFactors = FALSE
  ),
  data.frame(
    Time = time_points,
    Method = "Fine-Gray",
    Risk = risk_fg$P1[1, ],
    Disease = "ALL",
    stringsAsFactors = FALSE
  ),
  data.frame(
    Time = time_points,
    Method = "Fine-Gray",
    Risk = risk_fg$P1[2, ],
    Disease = "AML",
    stringsAsFactors = FALSE
  )
) %>% 
  dplyr::filter(Time <= 60)

@@

Figure \ref{fig:compAbsrisk} shows the cumulative incidence curves for all three models. As we can see, all three approaches agree quite well for AML; however, for ALL, there seems to be a difference of about 5% between the Fine-Gray curve and the curves estimated using case-base sampling and Cox regression. This difference does not appear to be significant: the curve from case-base sampling is contained within a 95% confidence band around the Fine-Gray absolute risk curve (figure not shown).

@@

In this article, we presented the R package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating risk functions using case-base sampling. Our package also provides several functions to produce graphical summaries of the data and the results. We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the case-base framework and the package through three case studies.

@@

As presented in Hanley & Miettinen [-@hanley2009fitting], case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the risk function. Accordingly, our package provides functions for each step. Moreover, the simple interface of the \code{fittingSmoothHazard} function resembles the \code{glm} interface. This interface should look familiar to new users. Our modular approach also provides a convenient way to extend our package for new sampling or fitting strategies.

In the case studies above, we compared the performance of case-base sampling with that of Cox regression and Fine-Gray models. In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula. By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that capture both the event type and time. Both approaches to modeling yield user-friendly code. However, in terms of output, both approaches differ significantly. Case-base sampling produces smooth hazards and smooth survival curves, whereas Cox regression and Fine-Gray models produce stepwise CIFs and never explicitly model the hazard function. Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves. However, the smooth nature of the output of \pkg{casebase} provides a more intuitive interpretation for consumers of these predictions.

<!-- In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling. -->

<!-- \begin{widetable} -->
<!-- \caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling} -->
<!-- \centering -->
<!-- \begin{tabular}[t]{llp{5cm}} -->
<!-- \toprule -->
<!-- Feature & Cox model & Case-base sampling\\ -->
<!-- \toprule -->
<!-- Model type & Semi-parametric & Fully parametric\\\midrule -->
<!-- Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\\midrule -->
<!-- Survival curve & Step function & Smooth-in-time curve\\\midrule -->
<!-- Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\\midrule -->
<!-- Model testing & Likelihood-ratio tests, score tests, & Use GLM framework\\ -->
<!--  & information criteria & (e.g.\ LRT, AIC, BIC)\\\midrule -->
<!-- \addlinespace -->
<!-- Competing risks & Aalen-Johansen & Cause-specific CIFs\\ -->
<!-- % Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\ -->
<!-- \bottomrule -->
<!-- \end{tabular} -->
<!-- \end{widetable} -->

@@

In future versions of the package, we also want to increase the complement of diagnostic and inferential tools that are currently available. For example, we would like to include more functions to compute calibration and discrimination statistics (e.g. AUC) for our models. Saarela and Arjas [-@saarela2015non] also describe how to obtain a posterior distribution for the AUC from their model. Their approach could potentially be included in \pkg{casebase}. Finally, we want to provide more flexibility in how the case-base sampling is performed. This could be achieved by adding a \code{hazard} argument to the function \code{sampleCaseBase}. In this way, users could specify their own sampling mechanism. For example, they could provide a hazard that gives sampling probabilities that are proportional to the cardiovascular disease event rate given by the Framingham score [@saarela2015non].

In conclusion, we presented the R package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth survival functions using the framework of generalized linear models. We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practitioners. The \pkg{casebase} package is freely available from the Comprehensive R Archive Network at \url{https://cran.r-project.org/package=casebase}. Interested users can visit \url{http://sahirbhatnagar.com/casebase/} for detailed package documentation and vignettes.
